---
title: "Image-Derived Color Data Wrangling in R"
output: html_document
---
## Overview

This repository demonstrates a reproducible workflow for processing and analyzing image-derived biological color data in R. The project includes simulated pixel-level datasets modeled after ImageJ exports and illustrates approaches for metadata parsing, channel integration, quantitative color metric calculation, and longitudinal data wrangling.

The workflow emphasizes reproducible analytical practices, including:
- structured project organization,
- automated file parsing,
- pixel-level data processing,
- feature engineering,
- and quantitative visualization.

Although inspired by biological signaling research, the methods demonstrated here are broadly applicable to image-based quantitative data analysis workflows.

Peak-red and peak-blue subsets represent localized regions with high RG or BY opponency values, respectively.


# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

set.seed(123)
```


## Packages

```{r packages}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  patchwork,
  here,
  data.table,
  knitr
)
```


## Helpers

```{r helpers}
normalize_name <- function(x) {
  x %>%
    as.character() %>%
    trimws() %>%
    tolower() %>%
    gsub("[[:space:]]+", "", .)
}

make_roi_id <- function(name, date, region) {
  paste(normalize_name(name), date, region, sep = "_")
}
```


# Color Data Wrangling
## File Parsing

Image-derived pixel files are automatically parsed to extract metadata, including channel identity, individual ID, collection date, and region of interest. The workflow expects ImageJ-style CSV exports containing X/Y pixel coordinates and channel-specific LMS values.

```{r file-parsing}
data.table::setDTthreads(1)
options(datatable.optimize = 1)

dir_in <- here("data", "pixelvalues")
out_path <- here("output", "color_metrics.csv")

dir.create(here("output"), recursive = TRUE, showWarnings = FALSE)

if (!dir.exists(dir_in)) {
  stop("Input directory not found: ", dir_in)
}

files <- list.files(dir_in, pattern = "(?i)\\.csv$", full.names = TRUE)
if (length(files) == 0) stop("No CSV files found in: ", dir_in)

file_info <- data.table(
  file = files,
  base = basename(files)
)
```


## Metadata Extraction and Validation

File names are parsed to identify channel-specific image exports and reconstruct region-of-interest identifiers. The workflow also checks whether each region has complete long-, medium-, and short-wavelength channel data before pixel-level processing.

```{r metadata-extraction}
file_info[, stem := sub("\\.[Cc][Ss][Vv]$", "", base)]
file_info[, chan_raw := str_extract(stem, "(?<=_)(lw|mw|sw)(?=_)")]
file_info[, chan := substr(chan_raw, 1, 1)]

if (anyNA(file_info$chan)) {
  bad <- file_info[is.na(chan), base]
  stop(
    "Couldn't detect lw/mw/sw channel token for some files. Examples:\n",
    paste(head(bad, 20), collapse = "\n")
  )
}

file_info[, roi_core := sub("_(lw|mw|sw)_", "_", stem)]

file_info %>%
  as_tibble() %>%
  select(base, chan_raw, chan, roi_core) %>%
  slice_head(n = 6) %>%
  knitr::kable(
    caption = "Example parsed file metadata"
  )
roi_core_info <- file_info[
  ,
  .(
    n_chans = uniqueN(chan),
    chans_present = paste(sort(unique(chan)), collapse = ",")
  ),
  by = roi_core
]

roi_core_info[
  ,
  date := str_extract(roi_core, "[0-9]{8}(?=_IMG)")
]

roi_core_info[
  ,
  `:=`(
    name = str_extract(roi_core, paste0("^.*?(?=", date, ")")),
    region = tolower(gsub("[0-9]+$", "", sub("^.*_", "", roi_core)))
  )
]

roi_core_info[
  ,
  period := case_when(
    date == "20230115" ~ "1",
    date == "20230302" ~ "2",
    date == "20230510" ~ "3",
    date == "20230718" ~ "4",
    date == "20230925" ~ "5",
    TRUE ~ NA_character_
  )
]

bad_parse <- roi_core_info[
  is.na(name) | is.na(date) | is.na(region) | is.na(period) |
    date == "" | region == "",
  roi_core
]

if (length(bad_parse) > 0) {
  stop(
    "Couldn't parse name/date/region/period for some roi_core values. Examples:\n",
    paste(head(bad_parse, 20), collapse = "\n")
  )
}

roi_core_info[, photo_id := sub("_[^_]+[0-9]+$", "", roi_core)]
roi_core_info[, male_id := normalize_name(name)]
roi_core_info[, id_main := make_roi_id(name, date, region)]
roi_core_info[, has_all_channels := n_chans == 3L]
```


## Image Summary and Channel Completeness

This section summarizes the number of photographs represented in the dataset and identifies regions of interest with complete long-, medium-, and short-wavelength channel data. Only complete regions are passed to the pixel-level processing workflow.

```{r image-summary-channel-completeness}
image_summary_df <- unique(roi_core_info[, .(male_id, photo_id)])

total_photos <- uniqueN(image_summary_df$photo_id)

photos_per_male <- image_summary_df[
  ,
  .(n_photos = .N),
  by = male_id
][order(male_id)]

photo_summary_stats <- photos_per_male[
  ,
  .(
    n_males = .N,
    mean_photos = mean(n_photos),
    sd_photos = sd(n_photos),
    min_photos = min(n_photos),
    max_photos = max(n_photos)
  )
]

photo_summary_stats
photos_per_male

missing_channel_cores <- roi_core_info[has_all_channels == FALSE, roi_core]
roi_cores_to_process <- roi_core_info[has_all_channels == TRUE, roi_core]

if (length(missing_channel_cores) > 0) {
  cat("Skipping ROI cores with incomplete channel sets:", length(missing_channel_cores), "\n")
}

cat("ROI cores selected for processing:", length(roi_cores_to_process), "\n")
```


## Pixel-Level Processing and Feature Extraction

For each complete region of interest, channel-specific pixel files are merged by X/Y coordinate. Low-signal pixels and extreme channel outliers are filtered before opponent-channel metrics are calculated. The workflow then summarizes full-region color values and extracts localized peak-red and peak-blue subsets using quantile thresholds.

```{r pixel-level-processing}
if (length(roi_cores_to_process) == 0) {
  stop("No complete ROI cores available for processing.")
}

file_info_new <- file_info[roi_core %in% roi_cores_to_process]

dir.create(here("output", "rawpixels"), recursive = TRUE, showWarnings = FALSE)

raw_rump_path     <- here("output", "rawpixels", "rawpixelvalues_rump.csv")
raw_peakred_path  <- here("output", "rawpixels", "rawpixelvalues_rump_peakred.csv")
raw_peakblue_path <- here("output", "rawpixels", "rawpixelvalues_rump_peakblue.csv")

suppressWarnings(file.remove(raw_rump_path, raw_peakred_path, raw_peakblue_path))

roi_results <- list()

first_rump <- TRUE
first_peakred <- TRUE
first_peakblue <- TRUE

aggregate_roi <- function(dt_in) {
  dt_in[
    ,
    .(
      l = mean(l, na.rm = TRUE),
      m = mean(m, na.rm = TRUE),
      s = mean(s, na.rm = TRUE),
      rg = mean(rg, na.rm = TRUE),
      by = mean(by, na.rm = TRUE),
      n_px = .N
    ),
    by = .(id, name, period, date, region, roi_core)
  ]
}

for (rc in roi_cores_to_process) {

  cat("Processing:", rc, "\n")

  fi <- file_info_new[roi_core == rc]


  # Read channel-specific pixel files and merge them into one table.
  dt_chan <- rbindlist(
    lapply(seq_len(nrow(fi)), function(i) {
      x <- fread(fi$file[i], select = c(1, 2, 3), showProgress = FALSE)
      setnames(x, c("X", "Y", "Value"))
      setnames(x, "Value", fi$chan[i])
      x
    }),
    use.names = TRUE,
    fill = TRUE
  )

  # Ensure all expected channel columns exist.
  for (cc in c("l", "m", "s")) {
    if (!cc %in% names(dt_chan)) dt_chan[, (cc) := NA_real_]
    dt_chan[[cc]] <- data.table::nafill(dt_chan[[cc]], type = "const", fill = 0)
  }

  # Collapse channel rows into one LMS value per pixel coordinate.
  dt_pix <- dt_chan[
    ,
    .(
      l = sum(l, na.rm = TRUE),
      m = sum(m, na.rm = TRUE),
      s = sum(s, na.rm = TRUE)
    ),
    by = .(X, Y)
  ]

  rm(dt_chan)
  gc()

  # Remove low-signal pixels.
  dt_pix <- dt_pix[l >= 0.0084 & m >= 0.0084 & s >= 0.0084]

  if (nrow(dt_pix) == 0) {
    cat("All candidate pixels removed for:", rc, "\n")
    next
  }

  # Attach metadata back to the pixel-level table.
  meta <- roi_core_info[roi_core == rc][1]

  dt_pix[
    ,
    `:=`(
      roi_core = rc,
      name = meta$name,
      date = meta$date,
      period = meta$period,
      region = meta$region,
      photo_id = meta$photo_id,
      id = meta$id_main
    )
  ]

  # Calculate simple opponent-channel color features.
  dt_pix[, rg := (l - m) / (l + m)]
  dt_pix[, by := s - ((l + m) / 2)]

  # Remove extreme channel outliers based on proportional LMS values.
  dt_pix[, total_lms := l + m + s]

  dt_pix[
    ,
    `:=`(
      l_prop = l / total_lms,
      m_prop = m / total_lms,
      s_prop = s / total_lms
    )
  ]

  q_m_high <- quantile(dt_pix$m_prop, 0.995, na.rm = TRUE)
  q_l_low  <- quantile(dt_pix$l_prop, 0.01,  na.rm = TRUE)
  q_s_low  <- quantile(dt_pix$s_prop, 0.005, na.rm = TRUE)

  dt_pix <- dt_pix[
    !(
      m_prop >= q_m_high &
        (l_prop <= q_l_low | s_prop <= q_s_low)
    )
  ]

  dt_pix[
    ,
    c("total_lms", "l_prop", "m_prop", "s_prop") := NULL
  ]

  if (nrow(dt_pix) == 0) next

  # Store full-region ROI summary.
  roi_results[[length(roi_results) + 1]] <- aggregate_roi(dt_pix)

  # For full-region samples, also extract localized peak-color subsets.
  if (meta$region == "rump") {

    fwrite(
      dt_pix[, .(id, name, date, region, X, Y, l, m, s)],
      raw_rump_path,
      append = !first_rump,
      col.names = first_rump
    )

    first_rump <- FALSE

    # Peak-red subset: upper 20% of red-green opponency values.
    q_rg_high <- quantile(dt_pix$rg, 0.80, na.rm = TRUE)

    dt_peakred <- dt_pix[rg >= q_rg_high]
    dt_peakred[, region := "rump_peakred"]
    dt_peakred[, id := make_roi_id(name, date, "rump_peakred")]

    roi_results[[length(roi_results) + 1]] <- aggregate_roi(dt_peakred)

    fwrite(
      dt_peakred[, .(id, name, date, region, X, Y, l, m, s)],
      raw_peakred_path,
      append = !first_peakred,
      col.names = first_peakred
    )

    first_peakred <- FALSE

    # Peak-blue subset: upper 20% of blue-yellow opponency values.
    q_by_high <- quantile(dt_pix$by, 0.80, na.rm = TRUE)

    dt_peakblue <- dt_pix[by >= q_by_high]
    dt_peakblue[, region := "rump_peakblue"]
    dt_peakblue[, id := make_roi_id(name, date, "rump_peakblue")]

    roi_results[[length(roi_results) + 1]] <- aggregate_roi(dt_peakblue)

    fwrite(
      dt_peakblue[, .(id, name, date, region, X, Y, l, m, s)],
      raw_peakblue_path,
      append = !first_peakblue,
      col.names = first_peakblue
    )

    first_peakblue <- FALSE

    rm(dt_peakred, dt_peakblue)
  }

  gc()
}

dt_roi_peak <- rbindlist(
  roi_results,
  use.names = TRUE,
  fill = TRUE
)

df_color_peak <- as_tibble(dt_roi_peak)

df_color_peak %>%
  count(region)

fwrite(dt_roi_peak, out_path)

cat("Saved ROI-level dataset:", out_path, "\n")
```


# Data Visualizations
## Opponent-Channel Distributions

The processed ROI-level dataset includes derived opponent-channel metrics representing red-green (RG) and blue-yellow (BY) chromatic variation.

```{r opponent-channel-distributions, fig.width=8, fig.height=4}
p_rg <- df_color_peak %>%
  ggplot(aes(rg)) +
  geom_histogram(
    bins = 30,
    fill = "#B22222",
    color = "white"
  ) +
  labs(
    title = "Distribution of Red-Green Opponency Values",
    x = "RG Opponency",
    y = "Count"
  ) +
  theme_minimal()

p_by <- df_color_peak %>%
  ggplot(aes(by)) +
  geom_histogram(
    bins = 30,
    fill = "#0072B2",
    color = "white"
  ) +
  labs(
    title = "Distribution of Blue-Yellow Opponency Values",
    x = "BY Opponency",
    y = "Count"
  ) +
  theme_minimal()

p_rg + p_by
```


## Region-Level Comparisons

Peak-red and peak-blue subsets are expected to differ in their opponent-channel characteristics relative to full-region summaries.

```{r region-comparison-boxplots, fig.width=8, fig.height=4}
p_region_rg <- df_color_peak %>%
  ggplot(aes(region, rg, fill = region)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(
    values = c(
      "rump" = "grey70",
      "rump_peakred" = "#B22222",
      "rump_peakblue" = "#F28E8C"
    )
  ) +
  labs(
    title = "RG Opponency Across ROI Types",
    x = "Region",
    y = "RG Opponency"
  ) +
  theme_minimal()

p_region_by <- df_color_peak %>%
  ggplot(aes(region, by, fill = region)) +
  geom_boxplot(show.legend = FALSE) +
  scale_fill_manual(
    values = c(
      "rump" = "grey70",
      "rump_peakred" = "#56B4E9",
      "rump_peakblue" = "#0072B2"
    )
  ) +
  labs(
    title = "BY Opponency Across ROI Types",
    x = "Region",
    y = "BY Opponency"
  ) +
  theme_minimal()

p_region_rg + p_region_by
```


## Example Pixel-Level Heatmap

The figure below shows an example spatial distribution of red-green opponency values across pixel coordinates from one processed region of interest.

```{r pixel-heatmap, fig.width=6, fig.height=5}
example_pixels <- fread(raw_rump_path)

example_pixels[, rg := (l - m) / (l + m)]

example_pixels <- example_pixels[
  id == unique(example_pixels$id)[1]
]

ggplot(example_pixels, aes(X, Y, fill = rg)) +
  geom_raster() +
  scale_fill_gradient(
    low = "white",
    high = "#B22222"
  ) +
  scale_y_reverse() +
  coord_fixed() +
  labs(
    title = "Example Spatial Distribution of RG Opponency",
    fill = "RG"
  ) +
  theme_minimal()
```
