---
title: "Longitudinal Behavioral Modeling with GLMMs in R"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Overview

This repository demonstrates a reproducible workflow for analyzing longitudinal behavioral data using generalized linear mixed models (GLMMs) in R. The project integrates simulated behavioral and quantitative color datasets to illustrate approaches for repeated-measures modeling, feature engineering, exploratory visualization, model diagnostics, and effect visualization.

The workflow emphasizes reproducible statistical practices, including:
- structured project organization,
- longitudinal data wrangling,
- mixed-effects modeling,
- model validation,
- and publication-style visualization.

Although inspired by behavioral ecology research, the analytical framework demonstrated here is broadly applicable to longitudinal and hierarchical data analysis problems across quantitative research domains.


# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

set.seed(123)
```


## Packages

```{r packages}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  glmmTMB,
  DHARMa,
  ggeffects,
  patchwork,
  here
)
```


## Load Data

Behavioral and quantitative color datasets are loaded and merged by individual identity and sampling period.

```{r load-data}
behavior_df <- read_csv(
  here("data", "behavior_data.csv"),
  show_col_types = FALSE
)

color_df <- read_csv(
  here("output", "color_metrics.csv"),
  show_col_types = FALSE
)

glimpse(behavior_df)
glimpse(color_df)
```

## Merge and Prepare Data

Predictor variables are standardized prior to modeling, and grouping variables are converted to factors for mixed-effects analysis.

```{r merge-and-prepare-data}
color_model_df <- color_df %>%
  filter(region == "rump") %>%
  transmute(
    id = name,
    period = as.character(period),
    rg,
    by,
    l,
    m,
    s,
    n_px
  )

df_models <- behavior_df %>%
  mutate(
    id = as.character(id),
    period = as.character(period)
  ) %>%
  left_join(
    color_model_df,
    by = c("id", "period")
  ) %>%
  mutate(
    id = factor(id),
    group = factor(group),
    period = factor(period),
    rank_z = as.numeric(scale(rank_prop)),
    rg_z = as.numeric(scale(rg)),
    by_z = as.numeric(scale(by))
  )

glimpse(df_models)
```

## Exploratory Visualization

Longitudinal trajectories and variable distributions are visualized prior to statistical modeling.

```{r exploratory-visualization, fig.width=8, fig.height=7}
p_longitudinal <- df_models %>%
  ggplot(aes(period, rg, group = id)) +
  geom_line(
    linewidth = 0.7,
    alpha = 0.5,
    color = "grey60"
  ) +
  geom_point(
    size = 1.3,
    alpha = 0.7,
    color = "#B22222"
  ) +
  geom_smooth(
    aes(group = 1),
    method = "loess",
    se = FALSE,
    color = "black",
    linewidth = 0.8
  ) +
  labs(
    title = "Longitudinal Changes in RG Opponency",
    x = "Sampling Period",
    y = "RG Opponency"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank()
  )

p_behavior <- df_models %>%
  ggplot(aes(reproductive_behavior_count)) +
  geom_histogram(
    bins = 12,
    fill = "#B22222",
    color = "white"
  ) +
  labs(
    title = "Distribution of Reproductive Behavior Counts",
    x = "Behavior Count",
    y = "Frequency"
  ) +
  theme_minimal()

p_longitudinal / p_behavior
```


## Generalized Linear Mixed Modeling

Generalized linear mixed models (GLMMs) were used to evaluate relationships between behavioral outcomes and quantitative signaling variables while accounting for repeated observations of individuals across sampling periods.

```{r glmm-model}
m_behavior <- glmmTMB(
  reproductive_behavior_count ~
    rank_z +
    rg_z +
    (1 | id),
  
  family = nbinom2,
  data = df_models
)

summary(m_behavior)
```

## Model Diagnostics

Simulated residual diagnostics were used to evaluate model fit and dispersion.

```{r model-diagnostics, fig.width=7, fig.height=5}
sim_behavior <- simulateResiduals(m_behavior)

plot(sim_behavior)
```


## Predicted Effects

Predicted marginal effects were estimated to visualize modeled relationships between signaling variables and behavioral outcomes.

```{r predicted-effects, fig.width=6, fig.height=5}
pred_rg <- ggpredict(
  m_behavior,
  terms = "rg_z"
)

ggplot(pred_rg, aes(x, predicted)) +
  geom_line(
    linewidth = 1,
    color = "#B22222"
  ) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.2,
    fill = "#B22222"
  ) +
  labs(
    title = "Predicted Effects of RG Opponency",
    x = "Standardized RG Opponency",
    y = "Predicted Behavioral Count"
  ) +
  theme_minimal()
```